---
title: "An automated procedure for phenology data analyses"
author: "Lea Rosa Maria Lipphardt"
date: "2024-11-19"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(repos = c(CRAN = "https://cran.rstudio.com/"))

```

## Introduction

This automated procedure streamlines the analysis of phenology data,
enabling more efficient estimation of key metrics such as hatching
dates, breeding success, and chick survival. The data is sourced from
the ACCESS database, specifically NEMO, and processed through a series
of R scripts.

The workflow is structured as follows:

1.  Import Data from the NEMO Database
2.  Combine Tables into a Single Data Frame
3.  Filter and Document Exceptions
4.  Estimate Hatching Dates
5.  Estimate Chick Survival
6.  Summarize Findings

This document is written in R Markdown, a versatile format for creating
HTML, PDF, and Word documents. For more details on R Markdown, visit
<http://rmarkdown.rstudio.com>.

To execute the R code, click the green arrow in the Script Editor
(Source pane). Press the Knit button to generate a final document that
integrates both narrative content and the output from your R code
chunks.

## 1. Import Data

Install the required packages and

```{r, results='hide', message=FALSE, warning=FALSE}

install.packages("odbc") # contains drivers to connect to a database
install.packages("DBI") # contains functions for interacting with the database

library(odbc)
library(DBI)
library(tidyverse)
library(dplyr)

```

1.1 Connect to the Access database using ODBC

```{r, results='hide', message=FALSE, warning=FALSE}
# Directs to the place the database is stored
dbname <- "N:/Midlertidig/Lea/NEMO/Spitsbergen_2024.mdb"

# Connect to the Access database using ODBC
con <- dbConnect(odbc::odbc(), 
                 .connection_string = paste0("Driver={Microsoft Access Driver (*.mdb, *.accdb)};DBQ=", dbname))

```

## 2. Query

2.1 Create dataframe using Structured query language (SQL)

```{r}
query <- "
SELECT s.spcENG, n.Locality, n.Area, n.NestNumber, v.VisitDate, nc.NestContentText, v.NestClutch
FROM (((Visits AS v
INNER JOIN NestContent AS nc ON v.NestStatus = nc.NestContentID)
INNER JOIN Nest AS n ON n.NestUniqueID = v.NestUniqueID)
INNER JOIN Species AS s ON n.Species = s.EUnr)
"
results <- dbGetQuery(con, query)

head(results)
str(results)


## Close the connection
dbDisconnect(con) # disconnect after your work is done to free resources
```

2.2 Control (no need to run this section)

```{r}
df_selected <- results
nrow(df_selected)

# Subset
#df_selected <- results[1:1000, ]

# Count the number of occurrences of each unique character in NestClutch
table(results$NestClutch)

# Convert df_selected into a tibble and summarize with dplyr
results %>%
  as_tibble() %>%
  count(NestClutch)
```

2.3 Convert dates into the right format and optional subset of data

```{r}
# Convert VisitDate to Date format if it's not already
df_selected$VisitDate <- as.Date(df_selected$VisitDate)

# Extract year from VisitDate (makes it easier to iterate)
df_selected$Year <- format(df_selected$VisitDate, "%Y")

# Subset the data for a specific year, specie, location
df_selected_year <- subset(df_selected, Year == "2024")
#df_selected_specie <- subset(df_selected_year, spcENG == "Black-legged kittiwake")
#df_selected_location <- subset(df_selected_specie, Locality == "Ossian Sarsfjellet")

```

## 3. Hatching dates

The hatching dates were estimated based on the last egg visit entry and
the first chick visit entry, a mean of the two registered dates.

3.1 Filtering steps and before getting hatching dates
Make a note for exceptions in a new column
```{r}
# Apply note to only egg/chick/empty entries and makes a note in a separate column 
filter_clutch_values <- function(df) {
  # Define valid clutch values for different categories
  egg_clutch_values <- c("1E", "2E", "3E", "1e", "2e", "3e", "1e,1E", "2e,2E", "3e,3E")
  chick_clutch_values <- c("1C", "2C", "3C", "1c", "2c", "3c", "1c,1C", "2c,2C", "3c,3C")
  empty_clutch_values <- c("")
  
  # Initialize the Note column with empty strings
  df$Note <- ""
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (year in unique(df$Year)) {
      # Subset the data for the current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$Year == year, ]
      
      # Remove rows with NA values in NestClutch
      subset_df <- subset_df[!is.na(subset_df$NestClutch), ]
      
      # Check conditions for E, C, or empty entries and append notes
      if (all(subset_df$NestClutch %in% egg_clutch_values)) {
        df$Note[df$NestNumber == nest & df$Year == year] <- "Only E-entries"
      } else if (all(subset_df$NestClutch %in% chick_clutch_values)) {
        df$Note[df$NestNumber == nest & df$Year == year] <- "Only C-entries"
      } else if (all(subset_df$NestClutch %in% empty_clutch_values)) {
        df$Note[df$NestNumber == nest & df$Year == year] <- "Only empty-entries"
      }
    }
  }
  
  return(df)
}

df_selected_year <- filter_clutch_values(df_selected_year)

head(df_selected_year)
```

3.2 Find last E and first C
```{r}
# Find for last visit with egg 
find_last_E <- function(df) {
  # Define the clutch values to search for
  clutch_values <- c("1E", "2E", "3E", "1e", "2e", "3e", "1e,1E", "3e,3E", "2e,3E", "1e,2E")
  
  # Create a new column `Last_E` initialized with NA
  df$Last_E <- as.Date(NA)
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (year in unique(df$Year)) {
      # Subset the data for current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$Year == year, ]
      
      # Find the last occurrence of any specified clutch value
      matching_clutches <- subset_df[subset_df$NestClutch %in% clutch_values, ]
      
      if (nrow(matching_clutches) > 0) {
        # Get the last matching row based on VisitDate
        last_entry <- matching_clutches[which.max(matching_clutches$VisitDate), ]
        
        # Assign the Last_E value to all rows for this NestNumber and Year
        df[df$NestNumber == nest & df$Year == year, "Last_E"] <- last_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Find for first visit with chick
find_first_C <- function(df) {
  # Define the clutch values to search for
  clutch_values <- c("1C,2E", "1E,1C", "2C", "1C", "1E,2C", "1c", "3C", "1C,1D", 
                     "2c", "3c", "1E,1c", "1e,1c", "3c,3C", "1E, 2c", "1c,2E", 
                     "1e,2c", "2E,2C", "1E,1C,1D", "1c,1C", "1e,2C", "1D,2C", "3e,3C")
  
  # Create a new column `First_C` initialized with NA
  df$First_C <- as.Date(NA)
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (year in unique(df$Year)) {
      # Subset the data for current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$Year == year, ]
      
      # Find the first occurrence of any specified clutch value
      matching_clutches <- subset_df[subset_df$NestClutch %in% clutch_values, ]
      
      if (nrow(matching_clutches) > 0) {
        # Get the first matching row based on VisitDate
        first_entry <- matching_clutches[which.min(matching_clutches$VisitDate), ]
        
        # Assign the First_C value to all rows for this NestNumber and Year
        df[df$NestNumber == nest & df$Year == year, "First_C"] <- first_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Find for last visit with chick 
find_last_C <- function(df) {
  # Define the clutch values to search for
  clutch_values <- c("1C,2E", "1E,1C", "2C", "1C", "1E,2C", "1c", "3C", "1C,1D", 
                     "2c", "3c", "1E,1c", "1e,1c", "3c,3C", "1E, 2c", "1c,2E", 
                     "1e,2c", "2E,2C", "1E,1C,1D", "1c,1C", "1e,2C", "1D,2C", "3e,3C")
  
  # Create a new column `Last_C` initialized with NA
  df$Last_C <- as.Date(NA)
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (year in unique(df$Year)) {
      # Subset the data for current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$Year == year, ]
      
      # Find the last occurrence of any specified clutch value
      matching_clutches <- subset_df[subset_df$NestClutch %in% clutch_values, ]
      
      if (nrow(matching_clutches) > 0) {
        # Get the last matching row based on VisitDate
        last_entry <- matching_clutches[which.max(matching_clutches$VisitDate), ]
        
        # Assign the Last_E value to all rows for this NestNumber and Year
        df[df$NestNumber == nest & df$Year == year, "Last_C"] <- last_entry$VisitDate
      }
    }
  }
  
  return(df)
}

# Get the last registered visit date 
find_last_visit <- function(df) {
  # Create a new column `Last_Visit` initialized with NA
  df$Last_Visit <- as.Date(NA)
  
  # Iterate over unique NestNumbers and Years
  for (nest in unique(df$NestNumber)) {
    for (year in unique(df$Year)) {
      # Subset the data for the current NestNumber and Year
      subset_df <- df[df$NestNumber == nest & df$Year == year, ]
      
      # Find the last visit based on VisitDate
      if (nrow(subset_df) > 0) {
        last_entry <- subset_df[which.max(subset_df$VisitDate), ]
        
        # Assign the Last_Visit value to all rows for this NestNumber and Year
        df[df$NestNumber == nest & df$Year == year, "Last_Visit"] <- last_entry$VisitDate
      }
    }
  }
  
  return(df)
}

phenology <- df_selected_year %>%
  find_last_E() %>%
  find_first_C() %>%
  find_last_C() %>%
  find_last_visit()

# View the result
head(phenology)
```

3.3 Estimate hatcing dates
```{r}
# Initialize new columns for hatching date and its accuracy
phenology$Hatching_date <- NA
phenology$Hatching_date_accuracy <- NA

# Get the hatching date and hatching date accuracy 
phenology <- phenology %>%
  mutate(
    # Calculate mean date for Hatching_date
    Hatching_date = if_else(
      !is.na(Last_E) & !is.na(First_C),
      as.Date((as.numeric(Last_E) + as.numeric(First_C)) / 2, origin = "1970-01-01"),
      NA_Date_
    ),
    
    # Calculate accuracy as the number of days divided by 2
    Hatching_date_accuracy = if_else(
      !is.na(Last_E) & !is.na(First_C),
      round(as.numeric(First_C - Last_E) / 2),
      NA_real_
    )
  )


# Select the columns of interest
phenology <- phenology %>%
  select(spcENG, Locality, Area, NestNumber, Year, Note, Last_E, First_C, Last_C, Last_Visit, Hatching_date, Hatching_date_accuracy)
head(phenology)
```
